Community biomass and species richness temporal trends reveal contrasting assembly and disassembly processes in natural food-webs

### 2024-08-08 paper/manuscript.Rmd
Rewrite intro + reorder appendix

```{r, echo = FALSE}
knitr::opts_chunk$set(
  cache = FALSE,
  collapse = TRUE,
  comment = "#>",
  #fig.dim = c(7, 7),
  fig.fullwidth = TRUE,
  fig.show = "hold",
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  results = TRUE,
  fig.path="figures/"
)
options(
  knitr.table.format = "latex",
  knitr.graphics.rel_path = FALSE
)
```

@@

# Abstract (240 / 250 words max)

Understanding food-web assembly processes is central to biological conservation
and maintenance of ecosystem functions. In particular, temporal changes in
community biomass and species richness are expected to significantly impact
food-web structure but empirical assessments are lacking so far. Here, we
investigate how temporal trends in species richness and community biomass affect
temporal trends in food-web structure. To do so, we used more than 6000 sampled
riverine fish community in 400 sites from french monitoring data. We inferred
food-web structure for each sampling event, using diet data and body size. Using
structural equation models, we estimated the effects of temporal changes in
community biomass and species richness on temporal changes in food-web
structure. We found that a positive effect of temporal trends in community
biomass on the temporal trends of connectance and average trophic level, a
result also found using spatial variations and a bioenergetic model. Temporal
trends in species richness had a positive effect on average trophic level, but
no effects on connectance. In space and in simulations, species richness had in
contrast a negative effect on average trophic level and connectance. In sum, our
study highlight that local temporal changes in biomass and species richness have
contrasting effects on food-web structure. It further reveals that temporal
changes and spatial variations in biomass lead to coherent food-web assembly
patterns, in contrast to species richness. Our study sheds a new light on
food-web assembly in natural settings.

# Significance (115 / 120 words max)

- Biodiversity changes = changes in food-web structure that are important for
  ecosystem functioning and stability.
- Unique combination of empirical data and theoretical simulations.
- Inform how community biomass and species richness declines result in an
  increase of the fragility of food-webs.

While the multidimensional nature of biodiversity changes has been clearly
emphasized, little is known on how the structure of food-web changes according to
changes in species richness and community biomass. By inferring food-web
structure based on food-diet data and body size ratios, we analyzed how food-web
structure in time according to concomitant changes in species richness and
community biomass. We identify that community biomass is the primary driver of
temporal changes in food-web structure, but that species richness has surprisingly
low effect. We further show that space-for-time leads to contrasting results for
species richness. Our results shed a new light on the assembly and disassembly
of food-web according to changes in biomass and in species richness.

@@

- Ideas:
  1. Biodiversity at risk; widespread heterogeneity in changes; ignores
     community struture
  2. Community structure is important for ecosystem functioning and robustness
     to perturbations
  3. Questions: assembly and disassembly processes of food-webs associated with
     biodiversity changes?; possible due to inference of food-web structure;
     Energetic constraints, redundancy and robustness; Validity of space for
     time hypothesis
  4. Presentation of the study data, theoretical models and extinction
     simulation: presentation of the hypothesis associated to the SEM

Why redundancy? Because not all links in the same way to the robustness of
food-webs

Understanding how biodiversity changes is a central question in ecology,
although challenging because biodiversity changes are multidimensional
[@mcgill_fifteen_2015; @blowes_geography_2019; @hillebrand_biodiversity_2018],
such as important changes in species composition can happen despite no changes
in species richness in average [@dornelas_assemblage_2014;
@blowes_geography_2019]. Yet species interactions are largely absent from
biodiversity changes. Although species interactions, described as ecological
networks, drive for a large part community composition, species coexistence, and
ecosystem functioning [@chesson_mechanisms_2000; @thebault_food-web_2003;
@oconnor_general_2017]. The structure of ecological networks can trigger or
impede perturbation propagations, i.e. secondary extinctions [@paine_note_1969;
@dunne_network_2002; @srinivasan_response_2007;
@stouffer_compartmentalization_2011]. Higher connectance, which measures the
degree of generalism in a network, is predicted to increase robustness to
secondary extinctions [@dunne_network_2002; @sole_complexity_2001] and studies
found that the presence of higher trophic levels is associated to a higher
temporal stability of community biomass in theoretical models and empirical data
[@shanafelt_stability_2018; @danet_species_2021]. Then, accounting for how the
structure of ecological networks changes over time according to other
biodiversity facets is crucial to understand biodiversity changes.

The availability of ecological network timeseries is scarce
[@kortsch_disentangling_2021], such as predictions about how the structure of
ecological networks change according to changes in species richness come from
spatial comparison of networks. Such studies showed that connectance is
negatively related to species richness in food-webs [reviewed in
@dunne_network_2006; @danet_species_2021], which reflects constraints on the
number of realizable species interactions or an increasing proportion of
specialists in more rich food-webs. Connectance has been shown to scale
positively or negatively with community biomass in theoretical food-webs
[@thebault_food-web_2003; @poisot_trophic_2013] while the average trophic level
was previously shown to scale negatively with species richness in empirical
food-webs [@danet_species_2021].

However, predicting how ecological networks will change in time according to
other biodiversity facets based on spatial comparison of food-webs
is valid if the assumption of space-for-time substitution is as
well. Space-for-time substitution supposes a homology between variations in
space and in time, i.e. that spatial sequences correspond to temporal sequences.
This method raised criticisms [@damgaard_critique_2019], with studies
showing analogous patterns in space and time [@blois_space_2013] or not
[@johnson_testing_2008; @adler_contrasting_2007]. Spatial turnover of species
composition, dispersal limitation, disturbance and stochastic processes have
been hypothesized to be responsible for the non-compliance of space-for-time
substitution [@adler_contrasting_2007; @blois_space_2013].

The increasing availability ecological temporal series and the development of
methods to reconstruct food-webs [@bonnaffe_comparison_2021;
@gravel_inferring_2013; @danet_species_2021] makes possible to overcome those
research gaps. We used a standardized monitoring of fish riverine communities on
the period 1995-2018 over France, documenting the individual size of fishes in
`r nb_st` sites during ten years or more. We reconstructed around `r nb_op`
food-webs based on allometric relationships and food diet data, and then
computed connectance, and average trophic level as well as community biomass and
species richness. We assessed how the local structure of food-webs changes over
time according to simultaneous changes in community biomass and species richness
over time and then if the relationships between changes in food-web structure,
community biomass and species richness, are the same over space and over time,
i.e. is the space-for-time substitution valid?

*Introduce hypothesis and the food-web metrics (connectance and average trophic
level)*

@@

Using a Bayesian hierarchical model to jointly estimate the temporal trends of
food-web structure at all sites, we found weak evidence of an overall decrease
in community biomass (CI80%:`r gbm` by decade; median [Credible Interval], Fig.
S1) but no evidence of an overall temporal increase in species richness
(CI80%: `r gr` by decade) across the `r nb_st` sites. We did not find evidence
of average temporal trends in connectance and average trophic level (CI80%:
`r gct` and `r gt` respectively), but an increase in redundancy (measured by the
proportion of redundant links, CI80%: `r gl`). However, we found a great
heterogeneity in the temporal trends estimated at the site level (community
biomass: from `r grbmn`% to `r grbmx`% per decade, species richness: `r grrhn`
to +`r grrhx`%, average trophic level: `r grtln` to +`r grtlx`, connectance:
`r grctn` to +`r grctx`, redundancy: `r grln` to +`r grlx`, Fig. S1) which was
well captured by the hierarchical model with conditional R2 going from
`r r2cond$log_bm_std` for community biomass to `r r2cond$log_rich_std` for
species richness (Table S1).

```{r}
sem_cap <- "(a) Structural equation models linking the temporal trends of the food-web
structure. The path widths are proportional to the standardized slope
coefficients. The standardized slope coefficients and their associated 95%
confidence intervals obtained by bootstrapping are displayed along the path. The
marginal R-squared of the linear models are displayed. (b to f) Direct bivariate
relationships derived from the structural equation model. The distribution of
the site temporal trends are displayed in Fig. S1. All the effects are reported
in Table S4."
```

@@

Using structural equation models to disentangle the direct and indirect effects
of temporal trends in fish species richness and community biomass on temporal
trends in food-web structure, we found a strong positive effect of the temporal
trends in species richness on community biomass ($r_\delta =$ `r rdtps$bm$r`,
Fig. \@ref(fig:sem) a and b, $r_\delta$: standardized slope coefficient, [95%
CI]), meaning that both temporal trends had the same direction in average. In
other word, food-webs that have experienced a decrease in species richness over
time have decreased in community biomass, and conversely, those that increased
in species richness have increased in community biomass.
Temporal trends in community biomass had in turn a positive effect on connectance,
average trophic level, and redundancy (resp. $r_\delta =$ `r rdtps$ct$bm`,
`r rdtps$tlvl$bm`, `r rdtps$red$bm`, Fig. \@ref(fig:sem)b, c, d), meaning that
food-web whose community biomass decreased over time have decreased in
connectance, average trophic level and redundancy, and
conversely, those that have experienced an increase in biomass have increased in
connectance, average trophic level and redundancy. We further found that the temporal
trends of community biomass were positively associated to the temporal trends of
both number and proportion of piscivorous species present (Fig. S3a), meaning
biomass trends are linked to the presence of high trophic level trophic
species.

We further tested if we find the same patterns using spatial variations in
food-web structure rather than temporal changes over time, i.e. if the
space-for-time hypothesis hold. To do so, we used the same dataset and same
structural equation model but with the median food-web structure per site
instead of the temporal trends. We additionally performed theoretical
simulations with a bioenergetic food-web model to further get theoretical
expectations about how food-web structure changes according to community biomass and
species richness variations (see Methods). As in temporal data, species richness
had a positive effect on community biomass both in empirical data and in
theoretical simulations (resp. $r_\delta =$ `r rdsp$bm$r` and `r rdm$bm$r`, Fig.
\@ref(fig:std) and S5). In both empirical data and theoretical simulations, community
biomass had a positive effect average trophic level (resp.
$r_\delta =$ `r rdsp$tlvl$bm` and `r rdm$tlvl$bm`), and redundancy (resp.
$r_\delta =$ `r rdsp$red$bm` and `r rdm$red$bm`), as in temporal analysis.
We found weak evidence of a positive effect of spatial variation in community
biomass on connectance in empirical data ($r_\delta =$ `r rdsp$ct$bm`) but
strong evidence in theoretical simulation ($r_\delta =$ `r rdm$ct$bm`, Fig. \@ref(fig:std)).

The effects of species richness were more complex, because species richness had
indirect effects on food-web structure through community biomass. Temporal
trends in species richness had a direct negative effect on connectance
($r_\delta =$ `r rdtps$ct$r`, Fig. \@ref(fig:sem)f), a positive
indirect effect through community biomass ($r_\delta =$ `r ridtps$ct$r`), so
that species richness temporal trends had a negative effect in total on temporal in
connectance ($r_\delta =$ `r rttps$ct$r`, Fig. \@ref(fig:std)).
In contrast, temporal trends in species had no direct effects on the temporal
trends in average trophic level ($r_\delta =$ `r rdtps$tlvl$r`, Fig.
\@ref(fig:sem)g), a positive indirect effect through community
biomass ($r_\delta =$ `r ridtps$tlvl$r`), which adds up to a positive total
effect of species richness temporal trends on average trophic level ($r_\delta
=$ `r rttps$tlvl$r`, Fig. \@ref(fig:std)). Species richness temporal trends had
direct and indirect positive effect on the temporal trends in redundancy (resp.
$r_\delta =$ `r rdtps$red$r`, `r ridtps$red$r`), which added up to a strong
positive effect ($r_\delta =$ `r rttps$red$r`). Furthermore, we found that
temporal trends of the proportion of piscivorous species was negatively
associated with the temporal trends of species richness (Fig. S3), meaning that
communities that increase in richness over time have an increase in the
proportion of non-piscivorous species over time and vice versa. It means
food-webs that have experienced declines in species richness over the last
decades have decreased in average trophic level and redundancy but increase in
connectance, and conversely, those that experienced increases in species
richness have increased in average trophic level, redundancy, but decreased in
connectance.

@@

We performed extinction simulations scenario on the metaweb, the food-web
including all the species and nodes found in the empirical data, to better
understand how species richness change affect food-web structure. When species
are removed in a sequential order beginning by the highest trophic levels,
connectance and average trophic level decreased as we removed species (red
lines, Fig. S2a). At the opposite, removing first low trophic level species first
increase connectance and average trophic level (green lines, Fig. S2a). In turn,
random species extinction sequence had little effect on food-web structure
(black lines, Fig. S2a).

```{r}
std_cap <- paste0(
  "Direct and total effects of community biomass and species richness on
  food-web structure. Standardized coefficients ($r_\\delta$) derived from the
  temporal, spatial, and bioenergetic model Structural Equation Models. Points
  and lines display the estimated effects and their confidence intervals at
  95\\%. Values are displayed in Table SXX, R-squared in Table SXX, and the
  visualisation of the bivariate relationships for the spatial and bioenergetic
  model SEMs are displayed in Figure SXX.")

@@

We found remarkable heterogeneity in community biomass and species richness
changes over the last decades in riverine fish communities, linked to changes in
food-web structure. Community biomass changes spanned from -`r grbmn`% to `r
grbmx`% per decade, species richness from `r grrhn`% to +`r grrhx`% per decade,
in line previous reports of the heterogeneity of local biodiversity changes in
terrestrial, marine and freshwater realms[@dornelas_assemblage_2014;
@blowes_geography_2019; @danet_past_2024]. In turn, connectance temporal changes
spanned from `r grctn` to +`r grctx` per decade, average trophic level, from
`r grtln` to +`r grtlx` per decade and the proportion of redundant trophic links
from `r grln` to +`r grlx` per decades. We questioned the effects of community
biomass and species richness changes over time on temporal changes in food-web
structure, that we also tested across space and with simulations using a
bioenergetic food-web model. We found positive effects of species richness
changes on changes in community biomass that were consistent across temporal and
spatial analysis and with the theoretical model. The effects of community
biomass changes on food-web structure were also consistent across analysis,
suggesting that loss of community biomass are linked to the loss of the trophic
species of high trophic level. However, changes in species richness had
contrasting effects on food-web structure depending on the type of analysis,
temporal or spatial and theoretical simulations. It suggests that food-web
structural changes induced by changes in species richness differs from the one
induced by the changes in community biomass, and that they imply different
processes across time and space. However, all analysis converged to indicate
that community biomass and species richness declines trigger a decrease in
food-web redundancy. In what follows, we discuss how changes in community
biomass and species richness changes over time and space are related to food-web
assembly and disassembly processes.

@@

We used a monitoring of freshwater fishes conducted all over France and
conducted by the Office Français de la Biodiversité (OFB, previously known as
ONEMA and AFB). The sampling protocols were standardized since 1995 and a
complete description of the sampling process can be found in
@danet_species_2021 (main text and appendix S2). The fishes are sampled using
electrofishing all across the streams in shallow streams and over the bank by
boat in deeper streams. All the fishes are counted and the bodysize of fishes are
recorded for a sample of them, enabling the estimate the distribution of
individual bodysize by species for each sampling [@bonnaffe_comparison_2021;
@danet_species_2021].

We selected the sites that have been monitored at least 10 years on the
1995-2018 period, with the same sampling protocol over time. We selected
one sampling a year by site. In each site, we ensured low temporal heterogeneity
in sampling period by selecting sampling that were performed in the same period,
leading to a mean standard deviation of sampled month in each site of three
weeks [@danet_species_2021]. Those steps leads to the selection of
`r nb_st` sites (Fig. \@ref(fig:matmet), Table S2).

@@

The performance of the model was assessed visually by plotting the fitted versus
the observed values (Fig. S4). We also visually inspected the Posterior
Integral Transform (PIT) and Conditional Predictive Ordinate (CPO) to assess the
quality of the model fit. We also computed the marginal ($R^2_m$)and conditional
($R^2_c$) R-squared of the models following previous formulations
[@nakagawa_general_2013; @lahuis_explained_2014]. In order to account for the
variability associated with the priors [@gelman_r-squared_2019], we computed the
variance of the posterior distribution  of each predicted values ($Var_{fit} =
\sigma^2(\hat{Y_i})$). We computed $R^2$ values for each predicted values, then
obtaining a $R^2$ distribution [@gelman_r-squared_2019]. We reported mean
marginal and conditional $R^2$ associated the 95% credible interval computed
using the Highest Posterior Density method.

@@

@@

We modeled the relationships among food-web structure, that is
connectance, average trophic level, the proportion of redundant trophic links, and
community biomass ($dB$) and species richness ($dR$) with Structural Equation
Models. We modeled species
richness as related to community biomass and food-web structure. In
turn community biomass was related to food-web structure. We computed one
structural equation model to assess the relationships among the temporal trends
and one for the relationships over the spatial gradient, we added a random effect
of the hydrographic basin on the intercept for the latter one.

>>>>

```{r, echo = FALSE}
knitr::opts_chunk$set(
  cache = FALSE,
  collapse = TRUE,
  comment = "#>",
  #fig.dim = c(7, 7),
  fig.fullwidth = TRUE,
  fig.show = "hold",
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  results = TRUE,
  fig.path = "figures/"
)
options(
  knitr.table.format = "latex",
  knitr.graphics.rel_path = FALSE
)
```

@@

# Abstract (250 / 250 words max)

While there are widespread evidence of biodiversity decline, little is known how
food-web disassemble in response. Such knowledge is crucial as food-web
structure determine ecosystem functioning and stability to perturbations.
However, assessing how changes in biodiversity affect changes in food-web
structure require the availability of food-web timeseries and to account for the
differential effects of different biodiversity facets on food-web structure.
Here, we used a standardized monitoring of riverine fish communities over the
last decades (1995-2018) in France, from which we inferred the food-web
structure for each sampling events using diet data and allometric scaling rules.
We estimated the temporal trends in community biomass, species richness, and
food-web structure using hierarchical Bayesian models. We assessed the effects
of community biomass and species richness temporal changes on temporal changes
in food-web structure: connectance, average trophic level and redundancy. We
further assess the generality and robustness of our findings by replicating the
analysis over space and using theoretical simulations with a bioenergetic model.
Our result suggest that food-webs that experienced a decline in community
biomass experienced conjointly a loss of species at upper trophic levels and
conversely. While species richness temporal trends had less clear effects on
food-web structure, mainly due through its indirect effects on community biomass
temporal trends. Both community biomass and species richness declines were
associated with a decrease in redundancy, raising concerns about robustness of
such communities to further perturbations. Our study highlight how biodiversity
changes affect food-web structure, raising concerns about consequences on
ecosystem functioning and robustness.

# Significance (109 / 120 words max)

Biodiversity temporal changes can lead to simultaneous changes in food-web
structure, which is turn can have consequences on ecosystem functioning and
stability. We provide empirical evidence that temporal changes in community
biomass and species richness are linked to changes contrasted assembly and
disassembly processes in food-webs in natural settings. We then show that loss
in both community biomass and species richness are related to a loss of
redundancy in the food-webs, which raise concerns about the robustness of such
communities to perturbations. We further added spatial and theoretical evidence
suggesting that our results are quite robust and open to assess how biodiversity
changes might affect ecosystem functioning and robustness.

@@

Biodiversity is experiencing massive declines due to continuously increasing
human pressures, which is expected to have striking consequences on ecosystem
functioning and fragility. It is increasing recognised that biodiversity
changes are widely heterogeneous spatially [@mcgill_fifteen_2015;
@blowes_geography_2019], such as species richness declines and increases can be
related to spatial variations in human pressures [@cardinale_is_2018;
@danet_past_2024]. It is increasing recognised that biodiversity changes alter
community structure [@dornelas_assemblage_2014; @hillebrand_biodiversity_2018;
@danet_past_2024], but so far, one key component of biodiversity, species
interactions, has been mostly ignored by studies assessing temporal changes in
biodiversity, despite the importance of biotic interactions for species
coexistence, ecosystem functioning and stability [@chesson_mechanisms_2000;
@thebault_food-web_2003; @oconnor_general_2017; @danet_species_2021]. A better
understanding of biodiversity changes will then need to account how biodiversity
changes alter the structure of ecological networks.

Yet, both species richness and food-web structure have a significant impact on
coexistence [@chesson_interaction_2008], ecosystem functioning
[@thebault_food-web_2003; @wang_biodiversity_2018], stability
[@danet_species_2021; @eschenbrenner_diversity_2023], and ecosystem robustness
to perturbations [@dunne_network_2002; @allesina_functional_2009;
@bodini_using_2009; @allesina_functional_2009]. Many predator species play a
keystone role by protecting preys from competitive exclusion
[@heithaus_predicting_2008; @johnson_rarity_2007; @estes_trophic_2011], thereby
maintaining species richness. Food-web with higher trophic levels also increase
ecosystem productivity [@thebault_food-web_2003; @wang_biodiversity_2018] and
community temporal stability [@danet_species_2021;
@eschenbrenner_diversity_2023]. Finally, the number and distribution of trophic
links in food-webs determine the robustness of communities to perturbations
[@dunne_network_2002; @allesina_functional_2009; @bodini_using_2009;
@allesina_functional_2009]. Accounting for how food-web assemble and disassemble
according to biodiversity changes is then crucial to evaluate the consequences
of biodiversity changes on ecosystem functioning and stability.

However, assessing the effects of biodiversity changes on food-web structure
requires the availability of multiple temporal timeseries of biodiversity and
food-webs structure, to disentangle the direct and indirect effects of
biodiversity changes on food-web structure, and to account for theories about
the multiple constraints on food-web assembly. We propose to monitor three
structural properties of food-webs and two common monitored biodiversity facets
(i.e. community biomass and species) that are grounded in food-web assembly,
ecosystem functioning, and stability research areas: connectance (i.e. degree of
generalism), weighted average trophic level (i.e. vertical location of biomass),
and redundancy (i.e. the proportion of redundant links). Their are pieces of
evidence to connect those food-web structural metrics to changes in ecosystem
functioning and robustness that we review before to present the temporal
timeseries that allows to assess them.

Food-web assembly theory provides predictions on how food-web structure should
change according to changes in community biomass and species richness. The
energetic paradigm and the trophic theory of biogeography
[@lindeman_trophic-dynamic_1942; @gravel_trophic_2011] stand that the presence
of top predators is conditional on the presence of preys in number enough that
it can provide enough energy from the bottom to the top to sustain high trophic
level species. In turn, it can that the presence of generalist predators, which is
positively associated to average trophic level and connectance, will be linked
more to community biomass changes than to species richness. In turn, connectance
and average trophic level are commonly reported to decrease with as species
richness increases [@dunne_network_2006; @danet_species_2021] because predators
are limited in the number of species they can consume and conversely preys are
limited in the number of predator they can sustain. Food-web structure has also
been linked to the robustness to species extinctions [@dunne_network_2002;
@allesina_who_2004; @bodini_using_2009]. Historically having being linked to
connectance, i.e. the number of links, more recent studies have proposed to
separate links that are functional or redundant [@allesina_who_2004;
@allesina_functional_2009; @bodini_using_2009]. The idea is that the loss of
functional links will trigger secondary extinctions because it suppresses
obligatory pathways from resources to consumers while the loss of redundant
links do not remove such obligatory pathways [@allesina_who_2004]. It means that
redundant trophic links enhance ecosystem robustness as they can be safely
removed without triggering secondary extinctions. Thereby we expected that
higher species richness results in higher redundancy in food-webs, meaning that
they create more independent pathways connecting basal producers to higher
trophic level. Finally, biodiversity and ecosystem functioning theory suggest
that in context of significant niche complementarity, community biomass should
be positively related to species richness [@wang_biodiversity_2018;
@schneider_animal_2016], such as it is necessary to account for the indirect
effects of species richness on food-web structure through community biomass.

Here, we report on a study that combine empirical assessment of food-web
structure and biodiversity temporal changes in natural settings, extinction
simulations in empirical food-webs, and theoretical simulations using
bioenergetic model. The availability of food-web timeseries used to be scarce
[@kortsch_disentangling_2021], but the increasing availability of biodiversity
timeseries and the development of methods to reconstruct food-webs
[@bonnaffe_comparison_2021; @gravel_inferring_2013; @danet_species_2021] makes
possible to overcome this limitation. We used a standardized monitoring of fish
riverine communities on the period 1995-2018 over France, documenting the
individual size of fishes in `r nb_st` sites during ten years or more (Table S1). We
reconstructed around `r nb_op` food-webs based on allometric relationships and
food diet data, and then computed connectance, and average trophic level
weighted by trophic species biomass as well as community biomass and species
richness (Fig. \@ref(fig:matmet)). This dataset and method has already been used
in previous studies [@danet_species_2021; @bonnaffe_comparison_2021;
@bonnaffe_interaction_2024]. Using hierarchical Bayesian models, we estimated
temporal trends in food-web structure, community biomass and richness at each
site [@danet_past_2024]. We then used structural equation models (SEM) to
assess the effects of community biomass and species richness temporal changes on
changes in food-web structure, while accounting for the effects of species
richness on community biomass. We then assessed if those effects are holding in
space, i.e. if the space-for-time substitution hypothesis hold
[@blois_space_2013; @damgaard_critique_2019; @adler_contrasting_2007]. We did so
by considering median values of food-web structure and community structure over
the timeseries at each site (Fig. \@ref(fig:matmet)). Finally, we ran
simulations using the bioenergetic model to assess if our results hold beyond
our empirical fish food-webs.

@@

Using a Bayesian hierarchical model to jointly estimate the temporal trends of
food-web structure at all sites, we found weak evidence of an overall decrease
in community biomass (CI80%:`r gbm` by decade; median [Credible Interval], Fig.
S1) but no evidence of an overall temporal increase in species richness
(CI80%: `r gr` by decade) across the `r nb_st` sites. We did not find evidence
of average temporal trends in connectance and average trophic level (CI80%:
`r gct` and `r gt` respectively), but an increase in redundancy (measured by the
proportion of redundant links, CI80%: `r gl`). However, we found a great
heterogeneity in the temporal trends estimated at the site level (community
biomass: from `r grbmn`% to `r grbmx`% per decade, species richness: `r grrhn`
to +`r grrhx`%, average trophic level: `r grtln` to +`r grtlx`, connectance:
`r grctn` to +`r grctx`, redundancy: `r grln` to +`r grlx`, Fig. S1) which was
well captured by the hierarchical model with conditional R2 going from
`r r2cond$log_bm_std` for community biomass to `r r2cond$log_rich_std` for
species richness (Table S2).

```{r}
sem_cap <- "(a) Structural equation models linking the temporal trends of the food-web
structure. The path widths are proportional to the standardized slope
coefficients. The standardized slope coefficients and their associated 95%
confidence intervals obtained by bootstrapping are displayed along the path. The
marginal R-squared of the linear models are displayed. (b to f) Direct bivariate
relationships derived from the structural equation model. The distribution of
the site temporal trends are displayed in Fig. S1. All the effects are reported
in Table S3."
```

@@

Using structural equation models to disentangle the direct and indirect effects
of temporal trends in fish species richness and community biomass on temporal
trends in food-web structure, we found a strong positive effect of the temporal
trends in species richness on community biomass ($r_\delta =$ `r rdtps$bm$r`,
Fig. \@ref(fig:sem) a and b, $r_\delta$: standardized slope coefficient, [95%
CI], Table S3), meaning that both temporal trends had the same direction in average. In
other word, food-webs that have experienced a decrease in species richness over
time have decreased in community biomass, and conversely, those that increased
in species richness have increased in community biomass.
Temporal trends in community biomass had in turn a positive effect on connectance,
average trophic level, and redundancy (resp. $r_\delta =$ `r rdtps$ct$bm`,
`r rdtps$tlvl$bm`, `r rdtps$red$bm`, Fig. \@ref(fig:sem)b, c, d), meaning that
food-web whose community biomass decreased over time have decreased in
connectance, average trophic level and redundancy, and
conversely, those that have experienced an increase in biomass have increased in
connectance, average trophic level and redundancy. We further found that the temporal
trends of community biomass were positively associated to the temporal trends of
both number and proportion of piscivorous species present (Fig. S2a), meaning
biomass trends are linked to the presence of high trophic level trophic
species.

We further tested if we find the same patterns using spatial variations in
food-web structure rather than temporal changes over time, i.e. if the
space-for-time hypothesis hold. To do so, we used the same dataset and same
structural equation model but with the median food-web structure per site
instead of the temporal trends. We additionally performed theoretical
simulations with a bioenergetic food-web model to further get theoretical
expectations about how food-web structure changes according to community biomass and
species richness variations (see Methods). As in temporal data, species richness
had a positive effect on community biomass both in empirical data and in
theoretical simulations (resp. $r_\delta =$ `r rdsp$bm$r` and `r rdm$bm$r`, Fig.
\@ref(fig:std) and S3). In both empirical data and theoretical simulations, community
biomass had a positive effect average trophic level (resp.
$r_\delta =$ `r rdsp$tlvl$bm` and `r rdm$tlvl$bm`), and redundancy (resp.
$r_\delta =$ `r rdsp$red$bm` and `r rdm$red$bm`), as in temporal analysis.
We found weak evidence of a positive effect of spatial variation in community
biomass on connectance in empirical data ($r_\delta =$ `r rdsp$ct$bm`) but
strong evidence in theoretical simulation ($r_\delta =$ `r rdm$ct$bm`, Fig.
\@ref(fig:std)).

The effects of species richness were more complex, because species richness had
indirect effects on food-web structure through community biomass. Temporal
trends in species richness had a direct negative effect on connectance
($r_\delta =$ `r rdtps$ct$r`, Fig. \@ref(fig:sem)f), a positive
indirect effect through community biomass ($r_\delta =$ `r ridtps$ct$r`), so
that species richness temporal trends had a negative effect in total on temporal in
connectance ($r_\delta =$ `r rttps$ct$r`, Fig. \@ref(fig:std)).
In contrast, temporal trends in species had no direct effects on the temporal
trends in average trophic level ($r_\delta =$ `r rdtps$tlvl$r`, Fig.
\@ref(fig:sem)g), a positive indirect effect through community
biomass ($r_\delta =$ `r ridtps$tlvl$r`), which adds up to a positive total
effect of species richness temporal trends on average trophic level ($r_\delta
=$ `r rttps$tlvl$r`, Fig. \@ref(fig:std)). Species richness temporal trends had
direct and indirect positive effect on the temporal trends in redundancy (resp.
$r_\delta =$ `r rdtps$red$r`, `r ridtps$red$r`), which added up to a strong
positive effect ($r_\delta =$ `r rttps$red$r`). Furthermore, we found that
temporal trends of the proportion of piscivorous species was negatively
associated with the temporal trends of species richness (Fig. S2), meaning that
communities that increase in richness over time have an increase in the
proportion of non-piscivorous species over time and vice versa. It means
food-webs that have experienced declines in species richness over the last
decades have decreased in average trophic level and redundancy but increase in
connectance, and conversely, those that experienced increases in species
richness have increased in average trophic level, redundancy, but decreased in
connectance.

@@

We performed extinction simulations scenario on the metaweb, the food-web
including all the species and nodes found in the empirical data, to better
understand how species richness change affect food-web structure. When species
are removed in a sequential order beginning by the highest trophic levels,
connectance and average trophic level decreased as we removed species (red
lines, Fig. S4a). At the opposite, removing first low trophic level species first
increase connectance and average trophic level (green lines, Fig. S4a). In turn,
random species extinction sequence had little effect on food-web structure
(black lines, Fig. S4a).

```{r}
std_cap <- paste0(
  "Direct and total effects of community biomass and species richness on
  food-web structure. Standardized coefficients ($r_\\delta$) derived from the
  temporal, spatial, and bioenergetic model Structural Equation Models. Points
  and lines display the estimated effects and their confidence intervals at
  95\\%. Values are displayed in Table S3, R-squared in Table S4, and the
  visualisation of the bivariate relationships for the spatial and bioenergetic
  model SEMs are displayed in Figure S3.")

@@

We found remarkable heterogeneity in community biomass and species richness
changes over the last decades in riverine fish communities, linked to changes in
food-web structure. Community biomass changes spanned from `r grbmn`% to
+`r grbmx`% per decade, species richness from `r grrhn`% to +`r grrhx`% per
decade, in line previous reports of the heterogeneity of local biodiversity
changes in terrestrial, marine and freshwater realms[@dornelas_assemblage_2014;
@blowes_geography_2019; @danet_past_2024]. In turn, connectance temporal changes
spanned from `r grctn` to +`r grctx` per decade, average trophic level, from
`r grtln` to +`r grtlx` per decade and the proportion of redundant trophic links
from `r grln` to +`r grlx` per decades. We questioned the effects of community
biomass and species richness changes over time on temporal changes in food-web
structure, that we also tested across space and with simulations using a
bioenergetic food-web model. We found positive effects of species richness
changes on changes in community biomass that were consistent across temporal and
spatial analysis and with the theoretical model. The effects of community
biomass changes on food-web structure were also consistent across analysis,
suggesting that loss of community biomass are linked to the loss of the trophic
species of high trophic level. However, changes in species richness had
contrasting effects on food-web structure depending on the type of analysis,
temporal or spatial and theoretical simulations. It suggests that food-web
structural changes induced by changes in species richness differs from the one
induced by the changes in community biomass, and that they imply different
processes across time and space. However, all analysis converged to indicate
that community biomass and species richness declines trigger a decrease in
food-web redundancy. In what follows, we discuss how changes in community
biomass and species richness changes over time and space are related to food-web
assembly and disassembly processes.

@@

We used a monitoring of freshwater fishes conducted all over France and
conducted by the Office Français de la Biodiversité (OFB, previously known as
ONEMA and AFB). The sampling protocols were standardized since 1995 and a
complete description of the sampling process can be found in
@danet_species_2021. The fishes are sampled using
electrofishing all across the streams in shallow streams and over the bank by
boat in deeper streams. All the fishes are counted and the bodysize of fishes are
recorded for a sample of them, enabling the estimate the distribution of
individual bodysize by species for each sampling [@bonnaffe_comparison_2021;
@danet_species_2021].

We selected the sites that have been monitored at least 10 years on the
1995-2018 period, with the same sampling protocol over time. We selected
one sampling a year by site. In each site, we ensured low temporal heterogeneity
in sampling period by selecting sampling that were performed in the same period,
leading to a mean standard deviation of sampled month in each site of three
weeks [@danet_species_2021]. Those steps leads to the selection of
`r nb_st` sites (Fig. \@ref(fig:matmet), Table S1).

@@

The performance of the model was assessed visually by plotting the fitted versus
the observed values (Fig. S5). We also visually inspected the Posterior
Integral Transform (PIT) and Conditional Predictive Ordinate (CPO) to assess the
quality of the model fit. We also computed the marginal ($R^2_m$)and conditional
($R^2_c$) R-squared of the models following previous formulations
[@nakagawa_general_2013; @lahuis_explained_2014]. In order to account for the
variability associated with the priors [@gelman_r-squared_2019], we computed the
variance of the posterior distribution  of each predicted values ($Var_{fit} =
\sigma^2(\hat{Y_i})$). We computed $R^2$ values for each predicted values, then
obtaining a $R^2$ distribution [@gelman_r-squared_2019]. We reported mean
marginal and conditional $R^2$ associated the 95% credible interval computed
using the Highest Posterior Density method.

@@

Overall, we found that standard deviation attributed to site effects in both
baseline and temporal trends in community structure were much higher than the
ones found for hydrographic basin (Table S5).

@@

We modelled the relationships among food-web structure, that is
connectance, average trophic level, the proportion of redundant trophic links, and
community biomass ($dB$) and species richness ($dR$) with Structural Equation
Models. We modelled species
richness as related to community biomass and food-web structure. In
turn community biomass was related to food-web structure. We computed one
structural equation model to assess the relationships among the temporal trends
and one for the relationships over the spatial gradient, we added a random effect
of the hydrographic basin on the intercept for the latter one.
